Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-09-23 analysis/paper/paper.Rmd
increase assessmentYrs to reduce impact of interannual variability

Analysis methods that can help make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. To be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates information about the impacts of landscape
change with available demographic data to reduce uncertainty in
population viability projections for boreal caribou [building on
@eacker_web-based_2019]. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
need for local monitoring varies with landscape condition, and to assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
original approach, and @mcloughlin_declines_2003,
@hervieux_widespread_2013, @eacker_web-based_2019, and @lamb2024 for
examples of application to caribou). Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022]. Groups typically also contain
uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
Some efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb2024], following broader trends in the
analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
population models, all available demographic or population data (e.g.
abundance surveys, vital rate estimates, trend data) are combined into a
single analysis that can permit novel parameter estimation, and improve
precision, and account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. These models are proving broadly useful for
rapid detection of population declines, evaluation of management
actions, and viability analysis of critically endangered species
[@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb2024] have not thus far included
demographic-disturbance relationships, and we are concerned that models
informed only by local demographic data could be used to inflate
uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou [building on @eacker_web-based_2019]. In our simple
proof-of-concept model, information is shared among populations using
priors derived from a published analysis of variation in demographic
rates among populations across Canada [@johnson_science_2020]. A more
rigorous statistical approach would be to integrate multiple populations
into a hierarchical Bayesian analysis of variation within and among
populations [@leasure_hierarchical_2019]; we have not done this because
owners have not authorized the use of the data for this purpose,
obtaining permission to use caribou demographic data is difficult, and
most managers only have access to their own local data. Despite
limitations, our simple Bayesian model represents an important
improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019]. We address this gap with a bias correction term
derived from a simple model of the recruitment survey observation
process.

@@

resultScenario = "S7"
resultSet ="cmult6ay3aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

We focus on analyzing survival of collared females (as in
@mcloughlin_declines_2003, @hervieux_widespread_2013 and
@eacker_web-based_2019), and the ratio of calves to cows in aerial
surveys of groups that include collared females [@nbckc_nbckc_2022],
because these remain the most widely available sources of demographic
information for boreal caribou populations [@johnson_science_2020].
Surveyed groups also contain uncollared animals so sample size is
typically larger for recruitment than for survival estimates. Errors
that can bias composition survey results include misidentifying young
bulls as cows or vice versa, and failing to detect calves
[@ellington_divergent_2020; @decesare_estimating_2012].

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. These
example population trajectories are informed by prior information about
the distribution of possible parameter values and outcomes from the
national analysis of demographic-disturbance relationships
[@johnson_science_2020] so we call it the "national" model. Our method
for selecting parameter values for each example trajectory yields a set
of alternative models (hypotheses) about the state of a particular
caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that our Bayesian model successfully
integrates local data with priors from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variability ($v_S,v_R$) by 75\\% in these examples. Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Following others [@eacker_web-based_2019; @lamb2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1, to allow sharing survival information across
years with a random effect [@eacker_web-based_2019,@lamb2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these
500 examples in which the Bayesian posterior mean population growth rate
and the true population growth rate are both greater than 0.99, or both
less than or equal to 0.99.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
when the forecast horizon is long (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)). Uncertainty is also higher in our high
anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
stochastic variation in population growth rates increase as populations
become very small (as in top right panel of Figure
\@ref(fig:examplePredictions)).

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are on average less
precise but not less accurate, which can happen when there is very small
prior probability that population is viable (orange band in the top
right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
lower levels of disturbance, local monitoring can reduce uncertainty,
but there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior (b) predictions from the Bayesian model for an example population trajectory. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, and collars are renewed each year.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort. When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}

@@

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus informed by a number of
lines of evidence that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.

There are a variety of possible actions intended to improve survival and
reproduction of populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models
in this paper do not include these actions, and cannot be used to
project outcomes for managed populations. Our intention is to reduce
uncertainty that can be used to delay conservation action; there is
nothing in our arguments, models or results that can or should be used
to discourage much needed monitoring of the effects of conservation
actions. Even when no additional monitoring is required for assessing
self-sustainability, there are a variety of reasons to continue
monitoring highly disturbed boreal caribou populations. Our point is
that action is also required, and lack of monitoring should not be used
as a reason to delay conservation measures for highly disturbed
populations.

@@

The main objective of Johnson et al's [-@johnson_science_2020;
@eccc_scientific_2011] national analysis was to inform a national
disturbance policy that applies to all ranges except SK1, and to inform
a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
hierarchical analysis [@leasure_hierarchical_2019] would be better
suited for estimating variation in demographic model parameters among
populations, and distinguishing that from variation among years.
However, data owners have not authorized the use of the data for that
purpose. In order to proceed with the available published information we
constrained our example trajectory simulations by assuming that
populations remain in their quantile, so populations that are lower than
or higher than average remain lower or higher than average over time.
The Bayesian model is not similarly constrained, and can accommodate
evidence of a shift from unexpectedly low to unexpectedly high
demographic rates over time, or vice versa. Switching from low to high
each year is unlikely, but there is evidence of variation in the slope
of demographic-disturbance relationships among populations
[@rudolph_demographic_2017] so it is important that the Bayesian model
allows for a greater range of possibilities than our more constrained
simulated example trajectories. Although our simulated trajectories do
not include all plausible possibilities we believe they are sufficient
to demonstrate how a Bayesian model can be used to integrate local data
with national priors, and to show how monitoring requirements depend on
landscape condition.

>>>>

Analysis methods that can help make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. To be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates information about the impacts of landscape 
change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
population viability projections. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
need for local monitoring varies with landscape condition, and to assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
original approach, and @mcloughlin_declines_2003,
@hervieux_widespread_2013, @eacker_web-based_2019, @lamb_effectiveness_2024 and @dalgarno_bboutools_2024 for
examples of application to caribou). Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain
uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the
analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
population models, available demographic or population data (e.g.
abundance surveys, vital rate estimates, trend data) are combined into a
single analysis that can permit novel parameter estimation, and improve
precision, and account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. These models are proving broadly useful for
rapid detection of population declines, evaluation of management
actions, and viability analysis of critically endangered species
[@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models
informed only by local demographic data could be used to inflate
uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou. In our simple proof-of-concept model, information is 
shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more
rigorous statistical approach would be to integrate multiple populations
into a hierarchical Bayesian analysis of variation within and among
populations [@leasure_hierarchical_2019]; we have not done this because
owners have not authorized the use of the data for this purpose,
obtaining permission to use caribou demographic data is difficult, and
most managers only have access to their own local data. Despite
limitations, our simple Bayesian model represents an important
improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation
process.

@@

resultScenario = "S10"
resultSet ="cmult6ay3aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

We focus on analyzing survival of collared females and the ratio of calves to cows in aerial
surveys of groups that include collared females [@nbckc_nbckc_2022],
because these remain the most widely available sources of demographic
information for boreal caribou populations [@johnson_science_2020].
Surveyed groups also contain uncollared animals so sample size is
typically larger for recruitment than for survival estimates. Errors
that can bias composition survey results include misidentifying young
bulls as cows or vice versa, and failing to detect calves
[@ellington_divergent_2020; @decesare_estimating_2012].

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. These
example population trajectories are informed by prior information about
the distribution of possible parameter values and outcomes from the
national analysis of demographic-disturbance relationships
[@johnson_science_2020] so we call it the "national" model. Our method
for selecting parameter values for each example trajectory yields a set
of alternative models (hypotheses) about the state of a particular
caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data with priors from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$). Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1 [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these
500 examples in which the Bayesian posterior mean population growth rate
and the true population growth rate over the full projection period are both greater than 0.99, or both less than or equal to 0.99.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
when the forecast horizon is long (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)), and when interannual variation is high (Figure
\@ref(fig:examplePredictions)). Uncertainty is also higher in our high
anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
stochastic variation in population growth rates increase as populations
become very small (as in top right panel of Figure
\@ref(fig:examplePredictions)).

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are on average less
precise but not less accurate, which can happen when there is very small
prior probability that population is viable (orange band in the top
right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
lower levels of disturbance, local monitoring can reduce uncertainty,
but in some circumstances there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). High interannual variation makes reducing uncertainty more difficult (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with high (b: $v_R=0.46,v_S=0.087$) and lower (c: $v_R=0.23,v_S=0.0435$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but in some circumstances there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}

@@

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus informed by a number of
lines of evidence that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.

There are a variety of possible actions intended to improve survival and
reproduction of populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models
in this paper do not include these actions, and cannot be used to
project outcomes for managed populations. Our intention is to reduce
uncertainty that can be used to delay conservation action; there is
nothing in our arguments, models or results that can or should be used
to discourage much needed monitoring of the effects of conservation
actions. Even when no additional monitoring is required for assessing
self-sustainability, there are a variety of reasons to continue
monitoring highly disturbed boreal caribou populations. The point we wish to emphasize is
that action is also required, and lack of monitoring should not be used
as a reason to delay conservation measures for highly disturbed
populations.

@@

The main objective of Johnson et al's [-@johnson_science_2020;
@eccc_scientific_2011] national analysis was to inform a national
disturbance policy that applies to all ranges except SK1, and to inform
a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
hierarchical analysis [@leasure_hierarchical_2019] would be better
suited for estimating variation in demographic model parameters among
populations, and distinguishing that from variation among years.
However, data owners have not authorized the use of the data for that
purpose. To develop a proof-of-concept model with available published information we
constrained our example trajectory simulations by assuming that
populations remain in their quantile, so populations that are lower than
or higher than average remain lower or higher than average over time.
The Bayesian model is not similarly constrained, and can accommodate
evidence of a shift from unexpectedly low to unexpectedly high
demographic rates over time, or vice versa. Switching from low to high
each year is unlikely, but there is evidence of variation in the slope
of demographic-disturbance relationships among populations
[@rudolph_demographic_2017] so it is important that the Bayesian model
allows for a greater range of possibilities than our more constrained
simulated example trajectories. Although our simulated trajectories do
not include all plausible possibilities we believe they are sufficient
to demonstrate how a Bayesian model can be used to integrate local data
with national priors, and to show how monitoring requirements depend on
landscape condition.
